Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES

### 2023-03-30 manuscript/paper.Rmd
change title and some little stuff

---
output: 
  stevetemplates::article:
    keep_tex: true
    fig_caption: true
    number_sections: yes
bibliography: bib/paper.bib
csl: bib/mee.csl
title: "Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES"
author:
- name: Victor Van der Meersch \footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr}, Isabelle Chuine
  affiliation: CEFE, Université de Montpellier, CNRS, EPHE, IRD, Montpellier, France
abstract: "1. Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models. Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and and their calibration methods \\textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}. \n\n 2. One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species \\textcolor{darkgreen}{compared to correlative SDMs}. We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data. This method is well established in some fields (robotics, aerospace research, ...), but has never been used, to our knowledge, in ecology, despite its ability to deal with very large space dimensions. Using tree species occurrence data across Europe, we adapted the CMA-ES algorithm to find appropriate values of model parameters. We estimated simultaneously 27 to 77 parameters of two process-based models simulating forest tree's ecophysiology for three species with varying range sizes and geographical distributions. \n\n 3. CMA-ES provided parameter estimates leading to better prediction of species distribution than parameter estimates based on experts knowledge. Our results also revealed that some model parameters and processes were strongly dependent, and different parameter combinations could therefore lead to high model accuracy. \n\n 4. \\textcolor{darkgreen}{We conclude that} CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data. Inverse modelling using CMA-ES is a powerful method to calibrate process-based parameters which can hardly be measured. However, the method does not warranty that parameter estimates are correct because of several sources of bias, similarly to correlative models, and expert knowledge is required to validate results. \n\n"
keywords: "calibration, evolutionary algorithm, cma-es, species distribution model, process-based model, trees"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
spacing: double
endnote: no
biblatex: true
link-citations: true

@@

Calibration (i.e. parameter setting and estimation) is a fundamental step in the modelling process. It allows the model to reproduce the reality with more or less success. The result of the calibration provides insights on the ability of the model to reproduce and explain reality (model predictive power). Calibration of complex models such as process-based SDMs is time-consuming, and modelers are often challenged by the dimension of the parameter space, the complexity of the possible correlations among parameters, and the scarcity of observed experimental data to calibrate them. Therefore, although the philosophy of such models is to measure parameters, statistical inference might be useful when data are not yet available to infer parameter values. Parameter inference can be achieved through several methods which have been developed in the last decades. \textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of *maximum likelihood estimation* and *Bayesian inference*, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding. One approach is to use numerical methods such as *simulation-based inference* or *approximate Bayesian computation*} [e.g. @Hartig2014]\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}     
Recently, an approach belonging to the evolutionary algorithm family, called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has been proposed [@Hansen2001]. One of the advantages of CMA-ES is its ability to cumulate information over iterations in order to adapt its own parameters (in particular the covariance matrix), which makes it more robust to noise. CMA-ES is especially performant for non-separable problems (i.e. when the model parameters are dependent) and large search space. This method has been successfully applied in various fields such as aerospace [e.g. @Collange2010], optics [e.g. @Gagne2008], and robotics [e.g. @Hill2020]. CMA-ES is acknowledged to be one of the most efficient approaches in continuous black-box optimization [@Hansen2010] but to our knowledge has never been used in ecology.

@@

Covariance Matrix Adaptation Evolution Strategy (CMA-ES) is widely accepted as a robust optimization algorithm for non-linear, non-convex, as well as non-separated optimization problems in continuous domain [@Hansen1996; @Hansen2001; @Hansen2006]. 
It is based on the principle of evolutionary biology, via recombination, mutation and selection of the most fit candidate solutions (i.e. parameter sets providing the best predictions). At each iteration:  
- $\lambda$ candidate solutions are evaluated, i.e. model runs $\lambda$ times with $\lambda$ different parameter sets and the objective function is evaluated  
- the best $\mu$ candidate solutionss are selected  
- the weighted mean candidate solution $m$ is computed (mean of the best $\mu$ parameter sets weighted by their objective function value)
- covariance matrix $C$ and step size $\sigma$ are updated (with information accumulated over several consecutive iterations)   
- new $\lambda$ candidate solutions are sampled in a normal distribution $\mathcal{N}(m, \sigma C)$, with both recombination (via the favorite solution $m$) and mutations (via the perturbations $\sigma C$)   
One of the strength of this approach lies in the combination of rank-$\mu$-update, where prior information from previous generations is exploited (mean of the previous covariance matrices, with a higher weight for recent generations), and cumulation, where correlations between generations are retained in an evolution path (sum of consecutive steps), to update the covariance matrix at each step [see @Hansen2016 for a detailed description of the algorithm].

@@

There are several issues regarding the process-based models we used which can impact their calibration and bias their parameter estimates. First, like any model, although they have a certain level of complexity, they are not a perfect representation of the reality, and their inherent structural errors increase the probability of finding parameter values that deviate from the true values of the underlying processes [see @Oberpriller2021]. Secondly, they do not necessarily include all the environmental factors at stake. For example, pedologic variable in PHENOFIT only involve the water holding capacity. In this model, other variables such as pH or soil texture are not considered. Thirdly, and more importantly, they are used here to represent the fundamental niche, and to estimate the potential distribution of the species using pedoclimatic variables. When calibrated against observed distributions, which represent the realized niche, they face the same issues as correlative models, and their parameter estimated can be distorted because compensating for processes not represented in the model (e.g. biotic interactions). In addition, in our case here, land use management probably also play an important role in shaping tree realized distribution, while not being addressed in the models. We included GBIF occurrence data, and especially as much as possible isolated native tree records outside forests, to help correct this problem, but it is impossible to be exhaustive at the spatial resolution of 0.1°. At this scale, local variations of soil characteristics and of competitive interactions among trees (e.g. along an altitudinal gradient) can also not be considered.

Finally, several authors advocate for process-based modeling approaches relying upon species response functions that are a priori defined [e.g. @Higgins2020]. However, the main limitation of such models is the data availability to infer their parameters [@Urban2016]. Expert parameterization is often long and arduous. One possible way to facilitate parameter value estimation would be to use inverse calibration, and we demonstrated here that CMA-ES can be a powerful optimizer to this end. For example, CMA-ES driven by species occurrence data could be used to calibrate submodels whose parameter values cannot be measured or are too hard to measure experimentally. However, when a structural correlation exists (i.e. trade-off between processes that are inherently present and interconnected in the model, as in the leaf unfolding submodel), inverse calibration might not provide the right parameter estimates. In such a case, expert knowledge, observations and measurements are necessary to determine *a posteriori* which estimates are the most realistic. This is possible in the case of process-based SDMs, and usually not feasable in the case of correlative SDMs. A combination of both expert and inverse calibrations might offer a new perspective for spreading the use of process-based models in predictive ecology, especially for climate change impact studies.

>>>>

---
output: 
  stevetemplates::article:
    keep_tex: true
    fig_caption: true
    number_sections: yes
bibliography: bib/paper.bib
csl: bib/mee.csl
title: "Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES"
author:
- name: Victor Van der Meersch \footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr}, Isabelle Chuine
  affiliation: CEFE, Université de Montpellier, CNRS, EPHE, IRD, Montpellier, France
abstract: "1. Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models. Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods \\textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}. \n\n 2. One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species \\textcolor{darkgreen}{compared to correlative SDMs}. We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data. This method is well established in some fields (robotics, aerospace research, ...), but has never been used, to our knowledge, in ecology, despite its ability to deal with very large space dimensions. Using tree species occurrence data across Europe, we adapted the CMA-ES algorithm to find appropriate values of model parameters. We estimated simultaneously 27 to 77 parameters of two process-based models simulating forest tree's ecophysiology for three species with varying range sizes and geographical distributions. \n\n 3. CMA-ES provided parameter estimates leading to better prediction of species distribution than parameter estimates based on experts knowledge. Our results also revealed that some model parameters and processes were strongly dependent, and different parameter combinations could therefore lead to high model accuracy. \n\n 4. \\textcolor{darkgreen}{We conclude that} CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data. Inverse modelling using CMA-ES is a powerful method to calibrate process-based parameters which can hardly be measured. However, the method does not warranty that parameter estimates are correct because of several sources of bias, similarly to correlative models, and expert knowledge is required to validate results. \n\n"
keywords: "calibration, evolutionary algorithm, cma-es, species distribution model, process-based model, trees"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
spacing: double
endnote: no
biblatex: true
link-citations: true

@@

Calibration (i.e. parameter setting and estimation) is a fundamental step in the modelling process. It allows the model to reproduce the reality with more or less success. The result of the calibration provides insights on the ability of the model to reproduce and explain reality (model predictive power). Calibration of complex models such as process-based SDMs is time-consuming, and modelers are often challenged by the dimension of the parameter space, the complexity of the possible correlations among parameters, and the scarcity of observed experimental data to calibrate them. Therefore, although the philosophy of such models is to measure parameters, statistical inference might be useful when data are not yet available to infer parameter values. Parameter inference can be achieved through several methods which have been developed in the last decades. \textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of \emph{maximum likelihood estimation} and \emph{Bayesian inference}, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding. One approach is to use numerical methods such as \emph{simulation-based inference} or \emph{approximate Bayesian computation} [e.g. @Hartig2014]\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}     
Recently, an approach belonging to the evolutionary algorithm family, called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has been proposed [@Hansen2001]. One of the advantages of CMA-ES is its ability to cumulate information over iterations in order to adapt its own parameters (in particular the covariance matrix), which makes it more robust to noise. CMA-ES is especially performant for non-separable problems (i.e. when the model parameters are dependent) and large search space. This method has been successfully applied in various fields such as aerospace [e.g. @Collange2010], optics [e.g. @Gagne2008], and robotics [e.g. @Hill2020]. CMA-ES is acknowledged to be one of the most efficient approaches in continuous black-box optimization [@Hansen2010] but to our knowledge has never been used in ecology.

@@

Covariance Matrix Adaptation Evolution Strategy (CMA-ES) is widely accepted as a robust optimization algorithm for non-linear, non-convex, as well as non-separated optimization problems in continuous domain [@Hansen1996; @Hansen2001; @Hansen2006]. 
It is based on the principle of evolutionary biology, via recombination, mutation and selection of the most fit candidate solutions (i.e. parameter sets providing the best predictions). At each iteration:  
- $\lambda$ candidate solutions are evaluated, i.e. model runs $\lambda$ times with $\lambda$ different parameter sets and the objective function is evaluated  
- the best $\mu$ candidate solutions are selected  
- the weighted mean candidate solution $m$ is computed (mean of the best $\mu$ parameter sets weighted by their objective function value)  
- covariance matrix $C$ and step size $\sigma$ are updated (with information accumulated over several consecutive iterations)   
- new $\lambda$ candidate solutions are sampled in a normal distribution $\mathcal{N}(m, \sigma C)$, with both recombination (via the favorite solution $m$) and mutations (via the perturbations $\sigma C$)   
One of the strengths of this approach lies in the combination of rank-$\mu$-update, where prior information from previous generations is exploited (mean of the previous covariance matrices, with a higher weight for recent generations), and cumulation, where correlations between generations are retained in an evolution path (sum of consecutive steps), to update the covariance matrix at each step [see @Hansen2016 for a detailed description of the algorithm].

@@

There are several issues regarding the process-based models we used which can impact their calibration and bias their parameter estimates. First, like any model, although they have a certain level of complexity, they are not a perfect representation of the reality, and their inherent structural errors increase the probability of finding parameter values that deviate from the true values of the underlying processes [see @Oberpriller2021]. Second, they do not necessarily include all the environmental factors at stake. For example, pedologic variable in PHENOFIT only involve the water holding capacity. In this model, other variables such as pH or soil texture are not considered. Third, and more importantly, they are used here to represent the fundamental niche, and to estimate the potential distribution of the species using pedoclimatic variables. When calibrated against observed distributions, which represent the realized niche, they face the same issues as correlative models, and their parameter estimated can be distorted because compensating for processes not represented in the model (e.g. biotic interactions). In addition, in our case here, land use management probably also play an important role in shaping tree realized distribution, while not being addressed in the models. We included GBIF occurrence data, and especially as much as possible isolated native tree records outside forests, to help correct this problem, but it is impossible to be exhaustive at the spatial resolution of 0.1°. At this scale, local variations of soil characteristics and of competitive interactions among trees (e.g. along an altitudinal gradient) can also not be considered.

Finally, several authors advocate for process-based modeling approaches relying upon species response functions that are a priori defined [e.g. @Higgins2020]. However, the main limitation of such models is the data availability to infer their parameters [@Urban2016]. Expert parameterization is often long and arduous. One possible way to facilitate parameter value estimation would be to use inverse calibration, and we demonstrated here that CMA-ES can be a powerful optimizer to this end. For example, CMA-ES driven by species occurrence data could be used to calibrate submodels whose parameter values cannot be measured or are too hard to measure experimentally. However, when a structural correlation exists (i.e. trade-off between processes that are inherently present and interconnected in the model, as in the leaf unfolding submodel), inverse calibration might not provide the right parameter estimates. In such a case, expert knowledge, observations and measurements are necessary to determine *a posteriori* which estimates are the most realistic. This is possible in the case of process-based SDMs, and usually not feasible in the case of correlative SDMs. A combination of both expert and inverse calibrations might offer a new perspective for spreading the use of process-based models in predictive ecology, especially for climate change impact studies.

### 2023-03-30 manuscript/paper.tex
change title and some little stuff

\noindent \begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Two main types of species distribution models are used to project
  species range shifts in future climatic conditions: correlative and
  process-based models. Although there is some continuity between these
  two types of models, they are fundamentally different in their
  hypotheses (statistical relationships \emph{vs} mechanistic
  relationships) and and their calibration methods
  \textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
\item
  One of the limitations to the use of process-based models is the
  difficulty to parameterize them for a large number of species
  \textcolor{darkgreen}{compared to correlative SDMs}. We investigated
  the feasibility of using an evolutionary algorithm (called covariance
  matrix adaptation evolution strategy, CMA-ES) to calibrate
  process-based models using species distribution data. This method is
  well established in some fields (robotics, aerospace research,
  \ldots), but has never been used, to our knowledge, in ecology,
  despite its ability to deal with very large space dimensions. Using
  tree species occurrence data across Europe, we adapted the CMA-ES
  algorithm to find appropriate values of model parameters. We estimated
  simultaneously 27 to 77 parameters of two process-based models
  simulating forest tree's ecophysiology for three species with varying
  range sizes and geographical distributions.
\item
  CMA-ES provided parameter estimates leading to better prediction of
  species distribution than parameter estimates based on experts
  knowledge. Our results also revealed that some model parameters and
  processes were strongly dependent, and different parameter
  combinations could therefore lead to high model accuracy.
\item
  \textcolor{darkgreen}{We conclude that} CMA-ES is an efficient
  state-of-the-art method to calibrate process-based models with a large
  number of parameters using species occurrence data. Inverse modelling
  using CMA-ES is a powerful method to calibrate process-based
  parameters which can hardly be measured. However, the method does not
  warranty that parameter estimates are correct because of several
  sources of bias, similarly to correlative models, and expert knowledge
  is required to validate results.
\end{enumerate}

@@

Calibration (i.e.~parameter setting and estimation) is a fundamental
step in the modelling process. It allows the model to reproduce the
reality with more or less success. The result of the calibration
provides insights on the ability of the model to reproduce and explain
reality (model predictive power). Calibration of complex models such as
process-based SDMs is time-consuming, and modelers are often challenged
by the dimension of the parameter space, the complexity of the possible
correlations among parameters, and the scarcity of observed experimental
data to calibrate them. Therefore, although the philosophy of such
models is to measure parameters, statistical inference might be useful
when data are not yet available to infer parameter values. Parameter
inference can be achieved through several methods which have been
developed in the last decades.
\textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of *maximum likelihood estimation* and *Bayesian inference*, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding. One approach is to use numerical methods such as *simulation-based inference* or *approximate Bayesian computation*}
(e.g. \protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
2014})\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data may be a reasonable alternative. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}\\
Recently, an approach belonging to the evolutionary algorithm family,
called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has
been proposed (\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier
2001}). One of the advantages of CMA-ES is its ability to cumulate
information over iterations in order to adapt its own parameters (in
particular the covariance matrix), which makes it more robust to noise.
CMA-ES is especially performant for non-separable problems (i.e.~when
the model parameters are dependent) and large search space. This method
has been successfully applied in various fields such as aerospace (e.g.
\protect\hyperlink{ref-Collange2010}{Collange \emph{et al.} 2010}),
optics (e.g. \protect\hyperlink{ref-Gagne2008}{Gagné \emph{et al.}
2008}), and robotics (e.g. \protect\hyperlink{ref-Hill2020}{Hill
\emph{et al.} 2020}). CMA-ES is acknowledged to be one of the most
efficient approaches in continuous black-box optimization
(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}) but to
our knowledge has never been used in ecology.

@@

Covariance Matrix Adaptation Evolution Strategy (CMA-ES) is widely
accepted as a robust optimization algorithm for non-linear, non-convex,
as well as non-separated optimization problems in continuous domain
(\protect\hyperlink{ref-Hansen1996}{Hansen \& Ostermeier 1996};
\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier 2001};
\protect\hyperlink{ref-Hansen2006}{Hansen 2006}). It is based on the
principle of evolutionary biology, via recombination, mutation and
selection of the most fit candidate solutions (i.e.~parameter sets
providing the best predictions). At each iteration:\\
- \(\lambda\) candidate solutions are evaluated, i.e.~model runs
\(\lambda\) times with \(\lambda\) different parameter sets and the
objective function is evaluated\\
- the best \(\mu\) candidate solutionss are selected\\
- the weighted mean candidate solution \(m\) is computed (mean of the
best \(\mu\) parameter sets weighted by their objective function value)
- covariance matrix \(C\) and step size \(\sigma\) are updated (with
information accumulated over several consecutive iterations)\\
- new \(\lambda\) candidate solutions are sampled in a normal
distribution \(\mathcal{N}(m, \sigma C)\), with both recombination (via
the favorite solution \(m\)) and mutations (via the perturbations
\(\sigma C\))\\
One of the strength of this approach lies in the combination of
rank-\(\mu\)-update, where prior information from previous generations
is exploited (mean of the previous covariance matrices, with a higher
weight for recent generations), and cumulation, where correlations
between generations are retained in an evolution path (sum of
consecutive steps), to update the covariance matrix at each step (see
\protect\hyperlink{ref-Hansen2016}{Hansen 2016} for a detailed
description of the algorithm).

@@

There are several issues regarding the process-based models we used
which can impact their calibration and bias their parameter estimates.
First, like any model, although they have a certain level of complexity,
they are not a perfect representation of the reality, and their inherent
structural errors increase the probability of finding parameter values
that deviate from the true values of the underlying processes (see
\protect\hyperlink{ref-Oberpriller2021}{Oberpriller \emph{et al.}
2021}). Secondly, they do not necessarily include all the environmental
factors at stake. For example, pedologic variable in PHENOFIT only
involve the water holding capacity. In this model, other variables such
as pH or soil texture are not considered. Thirdly, and more importantly,
they are used here to represent the fundamental niche, and to estimate
the potential distribution of the species using pedoclimatic variables.
When calibrated against observed distributions, which represent the
realized niche, they face the same issues as correlative models, and
their parameter estimated can be distorted because compensating for
processes not represented in the model (e.g.~biotic interactions). In
addition, in our case here, land use management probably also play an
important role in shaping tree realized distribution, while not being
addressed in the models. We included GBIF occurrence data, and
especially as much as possible isolated native tree records outside
forests, to help correct this problem, but it is impossible to be
exhaustive at the spatial resolution of 0.1°. At this scale, local
variations of soil characteristics and of competitive interactions among
trees (e.g.~along an altitudinal gradient) can also not be considered.

Finally, several authors advocate for process-based modeling approaches
relying upon species response functions that are a priori defined (e.g.
\protect\hyperlink{ref-Higgins2020}{Higgins \emph{et al.} 2020}).
However, the main limitation of such models is the data availability to
infer their parameters (\protect\hyperlink{ref-Urban2016}{Urban \emph{et
al.} 2016}). Expert parameterization is often long and arduous. One
possible way to facilitate parameter value estimation would be to use
inverse calibration, and we demonstrated here that CMA-ES can be a
powerful optimizer to this end. For example, CMA-ES driven by species
occurrence data could be used to calibrate submodels whose parameter
values cannot be measured or are too hard to measure experimentally.
However, when a structural correlation exists (i.e.~trade-off between
processes that are inherently present and interconnected in the model,
as in the leaf unfolding submodel), inverse calibration might not
provide the right parameter estimates. In such a case, expert knowledge,
observations and measurements are necessary to determine \emph{a
posteriori} which estimates are the most realistic. This is possible in
the case of process-based SDMs, and usually not feasable in the case of
correlative SDMs. A combination of both expert and inverse calibrations
might offer a new perspective for spreading the use of process-based
models in predictive ecology, especially for climate change impact
studies.

>>>>

\noindent \begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Two main types of species distribution models are used to project
  species range shifts in future climatic conditions: correlative and
  process-based models. Although there is some continuity between these
  two types of models, they are fundamentally different in their
  hypotheses (statistical relationships \emph{vs} mechanistic
  relationships) and their calibration methods
  \textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
\item
  One of the limitations to the use of process-based models is the
  difficulty to parameterize them for a large number of species
  \textcolor{darkgreen}{compared to correlative SDMs}. We investigated
  the feasibility of using an evolutionary algorithm (called covariance
  matrix adaptation evolution strategy, CMA-ES) to calibrate
  process-based models using species distribution data. This method is
  well established in some fields (robotics, aerospace research,
  \ldots), but has never been used, to our knowledge, in ecology,
  despite its ability to deal with very large space dimensions. Using
  tree species occurrence data across Europe, we adapted the CMA-ES
  algorithm to find appropriate values of model parameters. We estimated
  simultaneously 27 to 77 parameters of two process-based models
  simulating forest tree's ecophysiology for three species with varying
  range sizes and geographical distributions.
\item
  CMA-ES provided parameter estimates leading to better prediction of
  species distribution than parameter estimates based on experts
  knowledge. Our results also revealed that some model parameters and
  processes were strongly dependent, and different parameter
  combinations could therefore lead to high model accuracy.
\item
  \textcolor{darkgreen}{We conclude that} CMA-ES is an efficient
  state-of-the-art method to calibrate process-based models with a large
  number of parameters using species occurrence data. Inverse modelling
  using CMA-ES is a powerful method to calibrate process-based
  parameters which can hardly be measured. However, the method does not
  warranty that parameter estimates are correct because of several
  sources of bias, similarly to correlative models, and expert knowledge
  is required to validate results.
\end{enumerate}

@@

Calibration (i.e.~parameter setting and estimation) is a fundamental
step in the modelling process. It allows the model to reproduce the
reality with more or less success. The result of the calibration
provides insights on the ability of the model to reproduce and explain
reality (model predictive power). Calibration of complex models such as
process-based SDMs is time-consuming, and modelers are often challenged
by the dimension of the parameter space, the complexity of the possible
correlations among parameters, and the scarcity of observed experimental
data to calibrate them. Therefore, although the philosophy of such
models is to measure parameters, statistical inference might be useful
when data are not yet available to infer parameter values. Parameter
inference can be achieved through several methods which have been
developed in the last decades.
\textbackslash textcolor\{darkgreen\}\{Most of them fall into two
categories: informal and statistical calibration. On the one hand,
statistical calibration assumes a data-generating model, and a
likelihood function, which quantifies the probability of the observed
data given the model parameters, is expressed mathematically. This
likelihood is the foundation of \emph{maximum likelihood estimation} and
\emph{Bayesian inference}, two commonly used methods for parameter
estimation. In practice, deriving the likelihood function can be quite
challenging for complex models with many parameters and thresholding.
One approach is to use numerical methods such as
\emph{simulation-based inference} or
\emph{approximate Bayesian computation} (e.g.
\protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
2014})\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}\\
Recently, an approach belonging to the evolutionary algorithm family,
called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has
been proposed (\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier
2001}). One of the advantages of CMA-ES is its ability to cumulate
information over iterations in order to adapt its own parameters (in
particular the covariance matrix), which makes it more robust to noise.
CMA-ES is especially performant for non-separable problems (i.e.~when
the model parameters are dependent) and large search space. This method
has been successfully applied in various fields such as aerospace (e.g.
\protect\hyperlink{ref-Collange2010}{Collange \emph{et al.} 2010}),
optics (e.g. \protect\hyperlink{ref-Gagne2008}{Gagné \emph{et al.}
2008}), and robotics (e.g. \protect\hyperlink{ref-Hill2020}{Hill
\emph{et al.} 2020}). CMA-ES is acknowledged to be one of the most
efficient approaches in continuous black-box optimization
(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}) but to
our knowledge has never been used in ecology.

@@

Covariance Matrix Adaptation Evolution Strategy (CMA-ES) is widely
accepted as a robust optimization algorithm for non-linear, non-convex,
as well as non-separated optimization problems in continuous domain
(\protect\hyperlink{ref-Hansen1996}{Hansen \& Ostermeier 1996};
\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier 2001};
\protect\hyperlink{ref-Hansen2006}{Hansen 2006}). It is based on the
principle of evolutionary biology, via recombination, mutation and
selection of the most fit candidate solutions (i.e.~parameter sets
providing the best predictions). At each iteration:\\
- \(\lambda\) candidate solutions are evaluated, i.e.~model runs
\(\lambda\) times with \(\lambda\) different parameter sets and the
objective function is evaluated\\
- the best \(\mu\) candidate solutions are selected\\
- the weighted mean candidate solution \(m\) is computed (mean of the
best \(\mu\) parameter sets weighted by their objective function
value)\\
- covariance matrix \(C\) and step size \(\sigma\) are updated (with
information accumulated over several consecutive iterations)\\
- new \(\lambda\) candidate solutions are sampled in a normal
distribution \(\mathcal{N}(m, \sigma C)\), with both recombination (via
the favorite solution \(m\)) and mutations (via the perturbations
\(\sigma C\))\\
One of the strengths of this approach lies in the combination of
rank-\(\mu\)-update, where prior information from previous generations
is exploited (mean of the previous covariance matrices, with a higher
weight for recent generations), and cumulation, where correlations
between generations are retained in an evolution path (sum of
consecutive steps), to update the covariance matrix at each step (see
\protect\hyperlink{ref-Hansen2016}{Hansen 2016} for a detailed
description of the algorithm).

@@

There are several issues regarding the process-based models we used
which can impact their calibration and bias their parameter estimates.
First, like any model, although they have a certain level of complexity,
they are not a perfect representation of the reality, and their inherent
structural errors increase the probability of finding parameter values
that deviate from the true values of the underlying processes (see
\protect\hyperlink{ref-Oberpriller2021}{Oberpriller \emph{et al.}
2021}). Second, they do not necessarily include all the environmental
factors at stake. For example, pedologic variable in PHENOFIT only
involve the water holding capacity. In this model, other variables such
as pH or soil texture are not considered. Third, and more importantly,
they are used here to represent the fundamental niche, and to estimate
the potential distribution of the species using pedoclimatic variables.
When calibrated against observed distributions, which represent the
realized niche, they face the same issues as correlative models, and
their parameter estimated can be distorted because compensating for
processes not represented in the model (e.g.~biotic interactions). In
addition, in our case here, land use management probably also play an
important role in shaping tree realized distribution, while not being
addressed in the models. We included GBIF occurrence data, and
especially as much as possible isolated native tree records outside
forests, to help correct this problem, but it is impossible to be
exhaustive at the spatial resolution of 0.1°. At this scale, local
variations of soil characteristics and of competitive interactions among
trees (e.g.~along an altitudinal gradient) can also not be considered.

Finally, several authors advocate for process-based modeling approaches
relying upon species response functions that are a priori defined (e.g.
\protect\hyperlink{ref-Higgins2020}{Higgins \emph{et al.} 2020}).
However, the main limitation of such models is the data availability to
infer their parameters (\protect\hyperlink{ref-Urban2016}{Urban \emph{et
al.} 2016}). Expert parameterization is often long and arduous. One
possible way to facilitate parameter value estimation would be to use
inverse calibration, and we demonstrated here that CMA-ES can be a
powerful optimizer to this end. For example, CMA-ES driven by species
occurrence data could be used to calibrate submodels whose parameter
values cannot be measured or are too hard to measure experimentally.
However, when a structural correlation exists (i.e.~trade-off between
processes that are inherently present and interconnected in the model,
as in the leaf unfolding submodel), inverse calibration might not
provide the right parameter estimates. In such a case, expert knowledge,
observations and measurements are necessary to determine \emph{a
posteriori} which estimates are the most realistic. This is possible in
the case of process-based SDMs, and usually not feasible in the case of
correlative SDMs. A combination of both expert and inverse calibrations
might offer a new perspective for spreading the use of process-based
models in predictive ecology, especially for climate change impact
studies.
